---
title: "Lesson 5 : Advanced Spatial Point Patterns Analysis"
subtitle: "This study explores the concept of Local Colocation Quotients (Local CLQs)"
editor: visual
date:  6 Feb 2023
date-modified: "`r Sys.Date()`"
code-copy: true
execute: 
  echo: true
  eval: true
  warning: false
  error: false
website: 
    google-analytics: "G-VVFCV6MNHG"
format:
  html:
    code-overflow: wrap
    code-fold: true
    code-summary: "Show the code"
    css: styles.css
---

# **1. OVERVIEW**

This study explores the concept and application of Local Colocation Quotients (Local CLQs) with sfdep package.

There are three type CLQs :

-   Global

-   Pairwise

-   Local

## 1.1 Study Area

Taiwan

Refers to [EPSG.io](https://epsg.io/3829) for more details about CRS.

# 2. R PACKAGE REQUIRED

## **2.1 Load R Packages**

```{r}
#| code-fold: false
pacman::p_load(sfdep, sf, tmap, tidyverse) 
```

::: {.callout-tip collapse="true" appearance="simple" icon="false"}
## [**⇳ Highlights of the R Packages :**]{style="color:#3a9c88"}
:::

# **3. DATA PREPARATION**

## **3.1 Acquire Data Source**

This study will be based on trimmed data set provided with this lesson.

## **3.2 Import Data**

### **3.2.1 Import Attribute Data**

```{r}
studyArea <- st_read(dsn = "data",
                     layer = "study_area") %>%
  st_transform(crs = 3829)

```

```{r}
stores <- st_read(dsn = "data",
                  layer = "stores") %>%
  st_transform(crs = 3829)
```

### **3.2.2 Visualise sf Layers**

```{r}
tmap_mode("view")
tm_shape(studyArea) + 
  tm_polygons() +
  tm_shape(stores) +
  tm_dots(col = "Name",
          size = 0.01,
          border.col = "black",
          border.lwd = 0.5) +
  tm_view(set.zoom.limits = c(12,16))
```

```{r}
tmap_mode("plot")
```

# **4. LOCAL COLOCATION QUOTIENTS (LCLQ)**

## **4.1 Derive Arguments' Value**

### **4.1.1 Compute Neighbour List**

```{r}
nb <- include_self(st_knn
                   (st_geometry
                     (stores), 
                     6))

```

Remarks :

When using include_self( ) function, use even number to avoid 3-3 split scenario.

### **4.1.2 Derive Weight Matrix**

```{r}
wt <- st_kernel_weights(nb,
                        stores,
                        "gaussian",
                        adaptive = TRUE)
```

### **4.1.3 Derive Target Lists - A and B**

```{r}
FamilyMart <- stores %>%
  filter(Name == "Family Mart")

A <- FamilyMart$Name
```

```{r}
SevenEleven <- stores %>%
  filter(Name == "7-Eleven")
B <- SevenEleven$Name
```

## **4.2 Compute Local Simulations**

```{r}
LCLQ <- local_colocation(A,B,nb, wt, 49)
```

## **4.3 Visualise**

### **4.3.1 Combine Data Sets**

```{r}
LCLQ_stores <- cbind(stores, LCLQ)
```

Remarks :

-   Cannot use relational join such as left join as there is no unique identifier in LCLQ.

-   Do not sort LCLQ's data. Else, it won't be able to match back to stores.

-   Stores should be on the left side so that the geometry info in stores can be inherited. Else, it won't.

```{r}
tmap_mode("view")
tm_shape(studyArea) + 
  tm_polygons() +
  tm_shape(LCLQ_stores) +
  tm_dots(col = "X7.Eleven",
          size = 0.01,
          border.col = "black",
          border.lwd = 0.5) +
  tm_view(set.zoom.limits = c(12,16))

```
